---
subtitle: "Figure 3 supp"
title: "Violin plot for M-NeutroHigh vs other immune subgroup"
author: "Ilariya Tarasova"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    smart: yes
    toc: true
    toc_depth: 3
    code_folding: hide
    toc_float: true
    toc_collapsed: true
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE, message=FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
output_folder <- "../plots/"
if (!dir.exists(output_folder)){
    dir.create(output_folder, recursive = T)
}
```

# Request

> Sorin with CN spatial cluster, can you plot the violine plot to compare CN between the subgroup M-NeutroHigh vs L and another violine plot comparing M-NeutroHigh vs all the other subgroups combined (whole tissue only) – for maybe supp fig 3

# Read the data

the patient counts and new immune subgroups

```{r}
cn_res <- readRDS(file="../../Figure6/interim_data/out_CN6_Sorin_immune_only_new_kmean.rds")
#Percentage_Sorin_cohort_Tumour_Th <- readRDS("../../Figure2/interim_data/Percentage_Sorin_cohort_Tumour_MLLM_CD4CD14.rds")
#Sorin_cell_info <- readRDS(file="../../data/Sorin_cell_info.rds")
ImmuneSubgroups <- readRDS(file="../../data/Sorin_LUAD_immune_subgroups.rds")
patients_with_high_Neutr <- readRDS(file = "../../Figure3/interim_data/patients_with_high_Neutr_Sorin.rds")
```


## Add 

```{r}
#cn_res$InTumour <- Sorin_cell_info$InTumour[match(cn_res$cell.id, rownames(Sorin_cell_info))]

cn_res$Patient <- paste0("LUAD_", cn_res$Image.x)
```


# Summarise

```{r}
cn_counts <- cn_res %>% 
  group_by(Patient, CN.clusters) %>% 
  summarise(count = n(), .groups = "drop")

  clustered_counts <- cn_counts %>%
  filter(CN.clusters != -1) %>%
  group_by(Patient) %>%
  summarise(CN = sum(count), .groups = "drop")

# Step 2: Join back to original df
total_cn_counts <- cn_counts %>%
  filter(CN.clusters != -1) %>%
  left_join(clustered_counts, by = c("Patient"))

total_cn_counts$CN.clusters <- paste0("CN.", total_cn_counts$CN.clusters)
total_cn_counts_wide <- pivot_wider(total_cn_counts, names_from = "CN.clusters",
                                    values_from = "count", values_fill = 0)
```

## annotate with M-NeitroHigh


```{r}
total_cn_counts_wide$high_in_Neutr <- ifelse(total_cn_counts_wide$Patient %in% patients_with_high_Neutr,
                                              "yes", "no")
total_cn_counts_wide$immuneSubgroups <- ImmuneSubgroups$cluster[match(total_cn_counts_wide$Patient, ImmuneSubgroups$patient)]
total_cn_counts_wide$high_in_Neutr_subgr <- as.character(total_cn_counts_wide$immuneSubgroups)
# We are interested in patients from M subgroup that are high in Neutrophils
total_cn_counts_wide$high_in_Neutr_subgr <- ifelse(total_cn_counts_wide$high_in_Neutr_subgr == "M" &
                                                  total_cn_counts_wide$high_in_Neutr == "yes",
                                              "M_Neutr_high", total_cn_counts_wide$high_in_Neutr_subgr)
total_cn_counts_wide$high_in_Neutr_subgr <- ifelse(total_cn_counts_wide$high_in_Neutr_subgr == "M" &
                                                  total_cn_counts_wide$high_in_Neutr == "no",
                                              "M_Neutr_low", total_cn_counts_wide$high_in_Neutr_subgr)
# the order of the levels are in the way that the first one is a reference level
total_cn_counts_wide$high_in_Neutr_subgr <- factor(total_cn_counts_wide$high_in_Neutr_subgr,
                                                    levels = c("M_Neutr_high" ,"M_Neutr_low","LM", "L"))

# for the average
total_cn_counts_wide$only_Neutr_subgr <- total_cn_counts_wide$high_in_Neutr_subgr
total_cn_counts_wide$only_Neutr_subgr <- ifelse(total_cn_counts_wide$only_Neutr_subgr == "M_Neutr_high", 
                                                "M_Neutr_high", "other")
```

```{r}
df_long <- total_cn_counts_wide %>%
  pivot_longer(cols = starts_with("CN."), names_to = "CN_cluster", values_to = "count_in_cluster") %>%
  mutate(percent = 100 * count_in_cluster / CN)

# how many patients
length(unique(df_long$Patient))
# 416
```

# Visualisation

## Box plot

```{r}
immune_subgr_col <- c(M_Neutr_high = "#0868ac",
                      M_Neutr_low = "#43a2ca",
                      LM = "goldenrod1", L = "tomato1")

CN_cluster_col <- c(
  "CN.1" = "goldenrod1",
  "CN.2" = "red",
  "CN.3" = "cyan4",
  "CN.4" = "seagreen1",
  "CN.5" = "chocolate4",
  "CN.6" = "turquoise1"
)
matching <- c(
  "CN.1" = "CD4",
  "CN.2" = "B",
  "CN.3" = "MyeloT",
  "CN.4" = "Neutro",
  "CN.5" = "CD8",
  "CN.6" = "Macro"
)
df_long$CN_cluster_named <- matching[df_long$CN_cluster]

ggplot(df_long, aes(x=high_in_Neutr_subgr, y=percent, fill=CN_cluster)) + 
    geom_boxplot()+
  scale_fill_manual(values=CN_cluster_col, labels = matching)+
  theme_bw()
  #scale_fill_discrete(labels = matching)

ggplot(df_long, aes(x=CN_cluster, y=percent, fill=high_in_Neutr_subgr)) + 
    geom_boxplot()+
  scale_fill_manual(values=immune_subgr_col)+
  theme_bw()
```

## Volcano

### M vs L

```{r}
#average percentage per cluster and high_in_Neutr_subgr group
summary_df <- df_long %>%
  group_by(CN_cluster, high_in_Neutr_subgr) %>%
  summarise(mean_percent = mean(percent), .groups = "drop") %>%
  pivot_wider(names_from = high_in_Neutr_subgr, values_from = mean_percent) %>%
  mutate(log2FC = log2(M_Neutr_high / L)) 

p.values <- df_long %>%
  filter(high_in_Neutr_subgr %in% c("M_Neutr_high", "L")) %>% 
    group_by_at(c("CN_cluster")) %>%
    group_modify(~{
        return(data.frame(p.value=wilcox.test(percent ~ high_in_Neutr_subgr, .x, exact=FALSE)$p.value))
    })

summary_df$p.value <- p.values$p.value
```

```{r}
summary_df$CN_cluster_named <- matching[summary_df$CN_cluster]
ggplot(summary_df, aes(x=log2FC, y=-log10(p.value), colour = CN_cluster)) +
    geom_point(size=4, ) + 
    geom_vline(xintercept=0, linetype="dashed", colour = "grey60") + 
    geom_hline(yintercept = -log10(0.05), linetype="dashed", colour = "grey60") +
  labs(x = "log2FC(M_Neutr_high/L)", y = "-log10(p.value)",
       title = "Sorin, all data, spatial nbhd, M_Neutr_high vs L")+
    scale_color_manual(values=CN_cluster_col, labels = matching)+
    theme(
        panel.background = element_blank(),
        axis.line = element_line(colour = "black")
    )
ggsave(file.path(output_folder, paste0("volcano_M_Neutr_high_vs_L.png")))
```

### M vs the rest

```{r}
#average percentage per cluster and only_Neutr_subgr group
summary_df2 <- df_long %>%
  group_by(CN_cluster, only_Neutr_subgr) %>%
  summarise(mean_percent = mean(percent), .groups = "drop") %>%
  pivot_wider(names_from = only_Neutr_subgr, values_from = mean_percent) %>%
  mutate(log2FC = log2(M_Neutr_high / other)) 

p.values <- df_long %>%
    group_by_at(c("CN_cluster")) %>%
    group_modify(~{
        return(data.frame(p.value=wilcox.test(percent ~ only_Neutr_subgr, .x, exact=FALSE)$p.value))
    })

summary_df2$p.value <- p.values$p.value
```

```{r}
summary_df2$CN_cluster_named <- matching[summary_df2$CN_cluster]
ggplot(summary_df2, aes(x=log2FC, y=-log10(p.value), colour = CN_cluster)) +
    geom_point(size=4, ) + 
    geom_vline(xintercept=0, linetype="dashed", colour = "grey60") + 
    geom_hline(yintercept = -log10(0.05), linetype="dashed", colour = "grey60") +
  labs(x = "log2FC(M_Neutr_high/other)", y = "-log10(p.value)",
       title = "Sorin, all data, spatial nbhd, M_Neutr_high vs the rest")+
    scale_color_manual(values=CN_cluster_col, labels = matching)+
    theme(
        panel.background = element_blank(),
        axis.line = element_line(colour = "black")
    )
ggsave(file.path(output_folder, paste0("volcano_M_Neutr_high_vs_other.png")))
```